Personalised Map Matched Imputation: imputing missing data from smartphone location logs

### 2018-05-03 Thesis Manuscript/apaManu.tex
restructure and organise

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

@@

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            pdfauthor={},
            pdftitle={Handling missing data in smartphone location logs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Essential manuscript parts
  \title{Handling missing data in smartphone location logs}

@@

    \indent Word count: 2297
  }

@@

Why is human mobility important? How people move about in their
environment affects a wide range of outcomes, such as health, income and
social capital (Goodchild \& Janelle, 2010). Therefore it is
unsurprising that social scientists in numerous fields and even
policymakers are interested in human mobility measures. The most widely
administered personality questionnaires ask individuals to what extent
they agree with statements such as \enquote{I love large parties},
\enquote{I prefer going to the movies to watching videos at home} and
\enquote{I love to travel to places that I have never been before}
({\textbf{???}}). In economic research, the postal code of an
individual's home address is often used as a proxy for socioeconomic
status (e.g. Villanueva \& Aggarwal, 2013). Moreover, economists study
geographic labour mobility extensively (e.g. {\textbf{???}}). In fact,
the European Commission considers labour mobility important enough to
warrant its own directorate ({\textbf{???}}). The extensive use of
measures like these across different domains strongly suggests that
mobility metrics are linked to real world outcomes. Perhaps it is
unsurprising that behavioural researchers have found that mobility
measures can be used to predict academic performance (Wang, Harari, Hao,
Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz, \&
Odoms-Young, 2009) or even the onset of a depressive episode in bipolar
depression patients (Palmius et al., 2017). Indeed, there is an argument
to be made that perhaps psychologists have not been studying mobility
enough. When studying behavioural differences within individuals
behavioural scientists have often neglected the fact that individuals
vary not only over time but also over space. To fully understand
behaviour we must understand how behaviour can vary across environments.

Despite the importance of mobility measures, the majority of these
measures are obtained through the use of questionnaires (such as the
IPIP) or specifically through pen-and-paper travel diaries.
Questionnaires and travel diaries are have well-known methodological
flaws: they are burdensome to collect and rely on accurate
self-reporting. The burdensomeness stems from the fact that participants
must be explicitly asked to report on their movement patterns at
frequent intervals due to forgetfulness. This makes the data expensive
to collect and limits the extent of data collection in practice. In
addition, the frequent reporting duties of the participant may bias the
participants behaviour. Participant forgetfulness also limits the
accuracy of self reported measures. There is clear evidence that
participants are systematically biased when self-reporting mobility
measures. For instance, participants under-report the frequency of short
trips (Wolf, Oliveira, \& Thompson, 2003) and underestimate the duration
of regular commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017).
These obstacles can be overcome by using objective data on human
mobility.

Social scientists now have the unprecedented opportunity to easily
obtain objective data on human mobility from smartphones. Before
smartphones the only way to collect objective data on human mobility
involved giving participants an expensive professional-grade location
sensor and convincing them to take it with themselves at all times.
Barnett and Onnela (2016) points out that introducing a new device to
the participant's life may bias their behaviour. Moreover, collecting
data in such a way is costly, places a high burden on participants and
therefore the logs do not span a long time {[}Barnett and Onnela (2016).
Today millions of individuals carry smartphones with themselves every
day and do not need to be encouraged to do so by researchers.
Smartphones are equipped with a range of sensors that can be used to
track the location of the device at all times. These smartphones can
collect and store hundreds of location measurements a day. For instance,
Google Location History contains movement information on millions of
users, often spanning years (Location History, 2017). Moreover, recent
changes in EU regulations with regard to consumer data-portability
rights ensure that a willing participant should be able to easily share
this information with researchers at no cost to the either the
participant or the researchers (Commission, 2017). Taken together, this
means that researchers now have at their disposal the ability to easily
access the objective human mobility data of millions of individuals
spanning several years at little cost and without a significant burden
of the participants.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 1. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.}\label{fig:accuracyPlot2}
\end{figure}

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distance between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! This issue is also visualised in Figure
\ref{fig:palmiusVme1}. The problem is further complicated because
missing data and noisy measurements are related. Methods used by
researchers to reduce noise, such as throwing out inaccurate
measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
the missing data problem.

@@

A lack in methods for missing data imputation for human mobility
patterns does not imply there is not a vast literature on modelling
movement. The most widespread models are SSMs, therefore we shall detail
a few examples and subsequently argue that they are nonetheless unsuited
for long term human mobility logs. Ecologists have used SSMs to explain
how animals interact with their environment (Patterson, Thomas, Wilcox,
Ovaskainen, \& Matthiopoulos, 2008). These models can be quite complex.
Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. The extended Kalman filter is the de facto standard for GPS
navigation (Z. Chen \& Brown, 2013). The advantage of state space models
is that they are flexible, deal with measurement inaccuracy, include
information from different sources and can be used in real time.

@@

In the wider realm of spatiotemporal statistics there are numerous
missing data imputation methods. These often come from climate or
geological research and rely on spatiotemporal auto-correlations. For
instance, the CUTOFF method estimates missing values by incorporating
similar observed temporal information from the value's nearest spatial
neighbors (Feng, Nowak, O'Neill, \& Welsh, 2014 ). The authors
illustrate their example using rainfall data from gauging stations
across Australia. Similarly, Z. Zhang et al. (2017) use a variety of
machine learning methods to impute missing values. The example provided
relates to underground water data. Generally these models assume fixed
measurement stations (such as rainfall gauging stations). For this
reason they cannot be easily applied to missing mobility tracks without
significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns ({\textbf{???}}; Barnett \&
Onnela, 2016; Palmius et al., 2017 ). Importantly, none of them worked
with secondary logs. Nonetheless we will detail what they did as
informative examples. Palmius et al. (2017) deal with the measurement
inaccuracy of \(D\) in custom logs by removing from the data set all
unique low-accuracy \(a\) data points that had
\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
sample the data to a sample rate of 12 per hour using a median filter.
Moreover, Palmius et al. (2017) explain:

@@

This method can be extended to imputing the data based on temporally,
spatially or periodically close flights and pauses. In other words, for
a given missing period, the individual's mobility can be estimated based
on measured movements in that area, at that point in time or movements
in the last 24 hours (\emph{circadian proximity}).

On the other hand, ({\textbf{???}}) use what they call a Spatial
Temporal Semantic Neural Network (STS-NN) to predict future human
movement. While the authors are concerned with prediction and not
imputation, they devised a method called the Spatial Temporal Semantic
(STS) algorithm which converts raw measurements to machine learning
friendly discrete bins. Working with high-frequency measurements,
({\textbf{???}})'s method down-samples the raw data temporally and
map-matches the resulting bins to discrete points along pre-established
geographical features such as roads and highways. This minimises
measurement error and paves the way for applying machine learning
methods to human mobility problems.

In this section we have argued that there is a lack of established
practices to follow with respect to missing data in human mobility logs.
Moreover, extensively used spatiotemporal methods, such as state space
models (SSMs), are not well suited to deal with human mobility patterns
in secondary logs. Finally we discussed in detail three approaches which
deal explicitly with mobility patterns from custom or secondary logs
({\textbf{???}}; Barnett \& Onnela, 2016; Palmius et al., 2017 ).

@@

Following ({\textbf{???}})'s spatial-temporal-semantic (STS) feature
extraction algorithm our aim is to transform pause and path locations
into machine learning friendly discrete location sequences. There are
multiple ways of extracting such measurement clusters in the literature,
such as Spatio-Temporal Density-Based Spatial Clustering of Applications
with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
{[}(``ST-DBSCAN,'' 2007);{]}. We will focus on two methods which
explicitly with mobility patterns from unevenly sampled smartphone logs
(Barnett \& Onnela, 2016; Palmius et al., 2017 ). Both of these methods
pre-process the data and subsequently use two steps to extract pause
locations: first they extract pauses and their corresponding locations,
then they cluster pause locations based on spatial proximity. This
implementation of PMMI uses a stricter version of Barnett and Onnela
(2016)'s approach to extract pauses.

@@

({\textbf{???}})'s spatial-temporal-semantic (STS) feature extraction
algorithm uses map matching as a ground truth to assign noisy
measurements into discrete bins along roads. In other words, in addition
to the measured data they also use a geographic database that contains
information about the area in which the individual is (e.g.~where
precisely the roads are), and sort measurements into bins based on both
the measurement and the geographic data base. For example, if an
individual is measured as moving closely in parallel to a road A in an
area where there is no other parallel road, ({\textbf{???}})'s method
will assume that the individual is on the road A.

PMMI uses a similar logic, but without using any external geographic
database. The key modification in PMMI is that whilst ({\textbf{???}})
uses a map from outside the persons location logs, we use the total
location history of the individual to create a personalised map. This
map is subsequently used to bin measurements. This is feasible for two
reasons: humans tend to have repetitive movement habits and secondary
logs tend to be long. To put it in simpler terms, we consider each
measurement at \(D_x\) as a sample of \(Path_{ab}\), and by aggregating
many measurements we can use them to map out \(Path_{ab}\).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingdayBoaz5min} \caption{Missing data for the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. For this day there were several periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

For simplicity, we subsequently used a time period when the individual
was living in the Netherlands. This leaves us with 156,000 measurements
over a period of less than six months. Analyses as well as
implementations and adaptions Palmius et al. (2017)'s and Barnett and
Onnela (2016)'s model were performed using R and a multitude of other
statistical packages R (Version 3.4.3; R Core Team, 2017) and the
R-packages \emph{ggplot2} (Version 2.2.1; Wickham, 2009),
\emph{ggthemes} (Version 3.4.0; Arnold, 2017), \emph{knitr} (Version
1.20; Xie, 2015), and \emph{papaja} (Version 0.1.0.9709; Aust \& Barth,
2018). All the code is available on a public repository
({\textbf{???}}).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluating the performance of the map building and assigning
  functions.
\item
  Comparing the performance of PMMI using a) baseline models b)
  performance with randomly removed data in comparison to the
  aforementioned methods (Barnett \& Onnela, 2016; Palmius et al., 2017)
  and c) objective ground-truth data (public transportation
  time-stamps).
\end{enumerate}

@@

Before we can evaluate the accuracy of the imputations, it is essential
to evaluate how well noise in the data has been cleaned. Otherwise we
run the risk of over-fitting the model in the sense that we will measure
the extent to which an imputation method can correctly impute
measurement noise within \(D_t\) instead of true location \(G(t)\).

In order to evaluate the map building and binning we will first visually
evaluate the paths and pause locations. A visual evaluation of paths
superimposed on is an established way to heuristically check their
accuracy (e.g. ({\textbf{???}})). Then, let the average distance between
the actual measured point and the binned point be the \emph{deviation
distance} \(\delta_{\text{dev}}\). With respect to the deviation
distance \(\delta_{\text{dev}}\), we expect:

@@

\subsubsection{Imputation algorithm
performance}\label{imputation-algorithm-performance}

We will compare the performance of PMMI, Palmius et al. (2017) and
Barnett and Onnela (2016). In addition, we will also compute a naive
model, which simply imputes as the missing value the previous observed
value. The naive model will serve as a baseline model. To compare the
performance of these methods we will remove 25\% of measured time
intervals at random within a four week period. We will make our
comparisons for intervals of 5 minutes, 1 hour and 1 day. In other
words, we will remove 25\% of time intervals at random, while varying
the size of the time intervals removed.

@@

For the entire period examined period the we find a deviance of 40
meters and a median deviance of 15 meters. Around 69\% of the deviance
values are within their corresponding accuracy value, which is close to
the theoretical \(\frac{2}{3}\) value that is expected. Approximately
9\% of values were not taken into account when creating the bins,given
that their the accuracy \(a\) exceeded \(a_{\text{Path Lim}}\).

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

@@

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            pdfauthor={},
            pdftitle={Personalised Map Matched Imputation: imputing missing data from smartphone location logs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Essential manuscript parts
  \title{Personalised Map Matched Imputation: imputing missing data from
smartphone location logs}

@@

    \indent Word count: 8319
  }

@@

Why is human mobility important? Human mobility measures are quantified
metrics of how people move about in their environment. Human mobility
affects a wide range of outcomes, such as health, income and social
capital (Goodchild \& Janelle, 2010). The most widely administered
personality questionnaires ask individuals to what extent they agree
with statements such as \enquote{I love large parties}, \enquote{I
prefer going to the movies to watching videos at home} and \enquote{I
love to travel to places that I have never been before} (Goldberg et
al., 2006). At their root these questions are mobility measures. In
economic research, the postal code of an individual's home address is
often used as a proxy for socioeconomic status (e.g. Villanueva \&
Aggarwal, 2013). Economists are interested not only in where people
live, but also where they go to work: geographic labour mobility is
related to the income of individuals, the well-being of a country's
economy and even informs the policy of bodies such as the European
Comission (Tatsiramos, 2009). The extensive use of measures like these
across different domains in social science strongly suggests that
mobility metrics are linked to important real world outcomes. Perhaps it
is unsurprising that behavioural researchers have found that mobility
measures can be used to predict academic performance (R. Wang, Harari,
Hao, Zhou, \& Campbell, 2015), the incidence of obesity (Zenk, Schulz,
\& Odoms-Young, 2009) or even the onset of a depressive episode in
bipolar depression patients (Palmius et al., 2017). Indeed, there is an
argument to be made that perhaps psychologists have not been studying
mobility enough. When studying behavioural differences within
individuals behavioural scientists have often neglected the fact that
individuals vary not only over time but also over space. To fully
understand behaviour we must understand how behaviour can vary across
environments.

Despite the importance of mobility measures, the majority of these
metrics are obtained through the use of questionnaires (such as the
aforementioned personality questionnaire) or specifically through
pen-and-paper travel diaries. As a thought experiment we encourage the
reader to try to remember where exactly he or she was five weeks ago on
Saturday at midday. Most people find this sort of task very difficult
(if not impossible), which illustrates that questionnaires and travel
diaries are have well-known methodological flaws. This method of
collecting data is difficult and relies on accurate self-reporting.
Travel diaries are burdensome to collect because participants must be
explicitly asked to write down on their movement patterns at frequent
intervals due to human forgetfulness. This makes the data collection
expensive and limits the duration of data collection in practice. In
addition, the frequent reporting duties of the participant may bias the
participants behaviour. The limits of human cognition also limit the
accuracy of self-reported measures. There is evidence that participants
are systematically biased when self-reporting mobility measures. For
instance, participants under-report the frequency of short trips (Wolf,
Oliveira, \& Thompson, 2003) and underestimate the duration of regular
commutes (Delclòs-Alió, Marquet, \& Miralles-Guasch, 2017). These
obstacles can only be overcome by using objective data on human
mobility.

Social scientists now have the unprecedented opportunity to easily
obtain objective data on human mobility from smartphones. Before
smartphones the only way to collect objective data on human mobility
involved giving participants an expensive professional-grade location
sensor and convincing them to take it with themselves at all times.
Barnett and Onnela (2016) points out that introducing a new device to
the participant's life may bias their behaviour. Moreover, collecting
data in such a way is costly, places a high burden on participants and
therefore the logs do not span a long time (Barnett \& Onnela, 2016).
Today millions of individuals carry smartphones with themselves every
day and do not need to be encouraged to do so by researchers.
Smartphones are equipped with a range of sensors that can be used to
track the location of the device at all times. These smartphones can
collect and store hundreds of location measurements a day. For instance,
Google Location History contains movement information on millions of
users, often spanning years (Location History, 2017). Moreover, recent
changes in EU regulations with regard to consumer data-portability
rights ensure that a willing participant should be able to easily share
this information with researchers at no cost to the either the
participant or the researchers (Commission, 2017). Taken together, this
means that researchers now have at their disposal the ability to easily
access the objective human mobility data of millions of individuals
spanning several years at little cost and without a significant burden
of the participants.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{img/accuracyLocShift} \caption{Measures of user activity and measurement accuracy on February 15th 2017.The upper chart shows the distance from the next measured point in meters over the course of the day. The first peak corresponds to the first journey from the user's home to a gym around 8am. The second, smaller peak before 10 reflects a journey from the gym to the nearby lecture theatre. Both journeys can be seen in Figure 2. The large jump between journey 5 and 6 is measurement error. The lower chart shows the accuracy over the course of the day. The figure shows that measurement inaccuracy is sometimes related to the movement of the individual. Stationary accuracy varies depending on phone battery level, wifi connection and user phone use.}\label{fig:accuracyPlot2}
\end{figure}

Noisy data can lead to inaccurate conclusions if it is not accounted
for. Suppose we wish to calculate an individual's movement in a day. A
simple approach would be to calculate the sum of the distance between
each measurement. But if there is noise, the coordinates will vary even
though the individual is not moving. If the measurements are frequent
and noisy, we will calculate a lot of movement, even if the individual
did not move at all! This issue is also visualised in Figure
\ref{fig:palmiusVsMe1}. The problem is further complicated because
missing data and noisy measurements are related. Methods used by
researchers to reduce noise, such as throwing out inaccurate
measurements (e.g. Palmius et al., 2017), can exacerbate the severity of
the missing data problem.

@@

A lack in methods for missing data imputation for human mobility
patterns does not imply there is not a vast literature on modelling
movement. The most widespread models are SSMs, therefore we shall detail
a few examples and subsequently argue that they are nonetheless unsuited
for long term human mobility logs. Ecologists have used SSMs to explain
how animals interact with their environment (T. A. Patterson, Thomas,
Wilcox, Ovaskainen, \& Matthiopoulos, 2008). These models can be quite
complex. Preisler, Ager, Johnson, and Kie (2004) uses Markovian movement
processes to characterise the effect of roads, food patches and streams
on cyclical elk movements. The most well studied SSM is the Kalman
filter, which is the optimal algorithm for inferring linear Gaussian
systems. The extended Kalman filter is the de facto standard for GPS
navigation (Z. Chen \& Brown, 2013). The advantage of state space models
is that they are flexible, deal with measurement inaccuracy, include
information from different sources and can be used in real time.

@@

In the wider realm of spatiotemporal statistics there are numerous
missing data imputation methods. These often come from climate or
geological research and rely on spatiotemporal auto-correlations. For
instance, Feng, Nowak, O'Neill, and Welsh (2014) estimate missing values
by incorporating similar observed temporal information from the value's
nearest spatial neighbors. The authors illustrate their example using
rainfall data from gauging stations across Australia. Similarly, Z.
Zhang et al. (2017) use a variety of machine learning methods to impute
missing values. The example provided relates to underground water data.
Generally these models assume fixed measurement stations (such as
rainfall gauging stations). For this reason they cannot be easily
applied to missing mobility tracks without significant pre-processing.

On the other hand, a few researchers have explicitly attempted to impute
missing data from human mobility patterns (Barnett \& Onnela, 2016;
Palmius et al., 2017 ; @ Wu, Fu, Wang, Xiao, \& Fu, 2017). Importantly,
none of them worked with secondary logs. Nonetheless we will detail what
they did as informative examples. Palmius et al. (2017) deal with the
measurement inaccuracy of \(D\) in custom logs by removing from the data
set all unique low-accuracy \(a\) data points that had
\(\frac{d}{dt}D > 100 \frac{km}{h}\). Subsequently the researchers down
sample the data to a sample rate of 12 per hour using a median filter.
Moreover, Palmius et al. (2017) explain:

@@

This method can be extended to imputing the data based on temporally,
spatially or periodically close flights and pauses. In other words, for
a given missing period, the individual's mobility can be estimated based
on measured movements in that area, at that point in time or movements
in the last 24 hours.

It is also worth mentioning that there is a body of work that
incorporates road features and other background information to generate
predictions or imputations. For instance, Liao, Patterson, Fox, and
Kautz (2007) use hierarchical Markov models in combination with
knowledge about the transportation system to bridge the gap between raw
GPS sensor measurements and high level information such as a user's
destination and mode of transportation. This can be then used to impute
or predict missing steps. Along similar lines Wu et al. (2017) use what
they call a Spatial Temporal Semantic Neural Network (STS-NN) to predict
future human movement. While the authors are concerned with prediction
and not imputation, they devised a method called the Spatial Temporal
Semantic (STS) algorithm which converts raw measurements to machine
learning friendly discrete bins. Working with high-frequency
measurements, Wu et al. (2017)'s method down-samples the raw data
temporally and map-matches the resulting bins to discrete points along
pre-established geographical features such as roads and highways. This
minimises measurement error and paves the way for applying machine
learning methods to human mobility problems. Subsequently we will focus
on methods which do not incorporate external information as they are
more generally applicable.

In this section we have argued that there is a lack of established
practices to follow with respect to missing data in human mobility logs.
Moreover, extensively used spatiotemporal methods, such as state space
models (SSMs), are not well suited to deal with human mobility patterns
in secondary logs. Finally we discussed in detail three approaches which
deal explicitly with mobility patterns from custom or secondary logs
(Barnett \& Onnela, 2016; Palmius et al., 2017 ; Wu et al., 2017).

@@

Following Wu et al. (2017)'s spatial-temporal-semantic (STS) feature
extraction algorithm our aim is to transform pause and path locations
into machine learning friendly discrete location sequences. There are
multiple ways of extracting such measurement clusters in the literature,
such as Spatio-Temporal Density-Based Spatial Clustering of Applications
with Noise (ST-DBSCAN) and sequence oriented clustering (SOC)
(``ST-DBSCAN,'' 2007). We will focus on two methods which explicitly
with mobility patterns from unevenly sampled smartphone logs (Barnett \&
Onnela, 2016; Palmius et al., 2017 ). Both of these methods pre-process
the data and subsequently use two steps to extract pause locations:
first they extract pauses and their corresponding locations, then they
cluster pause locations based on spatial proximity. This implementation
of PMMI uses a stricter version of Barnett and Onnela (2016)'s approach
to extract pauses.

@@

Wu et al. (2017)'s spatial-temporal-semantic (STS) feature extraction
algorithm uses map matching as a ground truth to assign noisy
measurements into discrete bins along roads. In other words, in addition
to the measured data they also use a geographic database that contains
information about the area in which the individual is (e.g.~where
precisely the roads are), and sort measurements into bins based on both
the measurement and the geographic data base. For example, if an
individual is measured as moving closely in parallel to a road A in an
area where there is no other parallel road, Wu et al. (2017)'s method
will assume that the individual is on the road A.

PMMI uses a similar logic, but without using any external geographic
database. The key modification in PMMI is that whilst Wu et al. (2017)
uses a map from outside the persons location logs, we use the total
location history of the individual to create a personalised map. This
map is subsequently used to bin measurements. This is feasible for two
reasons: humans tend to have repetitive movement habits and secondary
logs tend to be long. To put it in simpler terms, we consider each
measurement at \(D_x\) as a sample of \(Path_{ab}\), and by aggregating
many measurements we can use them to map out \(Path_{ab}\).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{img/missingdayBoaz5min} \caption{Missing data for the entire duration of the log. The x-axis denotes time, the y-axis shows how many measurements are made and each point is a five minute window. The entire log contains several long periods with no information. These points are filled with red and lie on the x-axis.}\label{fig:measurementsPerDay}
\end{figure}

For simplicity, we subsequently used a time period when the individual
was living in the Netherlands. This subset contains 156,000 measurements
over a period of less than six months.

Palmius et al. (2017)'s algorithm was implemented in R based on the
original MATLAB code and pseudocode kindly provided by the author.
Barnett and Onnela (2016)'s method was slightly adapted in R to fit the
the data structure from the original R code provided by the author. When
executing their models we used the same parameters as the authors did.
To represent Barnett and Onnela (2016)'s model we used the variant where
movements were sampled based on spatial proximity.

All analyses were performed using R (Version 3.4.3; R Core Team,
2017)\footnote{We, furthermore, used the R-packages \emph{bindrcpp}
  (Version 0.2; Müller, 2017), \emph{dbplyr} (Version 1.2.0; Wickham \&
  Ruiz, 2018), \emph{dplyr} (Version 0.7.4; Wickham, Francois, Henry, \&
  Müller, 2017), \emph{geosphere} (Version 1.5.7; Hijmans, 2017a),
  \emph{ggplot2} (Version 2.2.1; Wickham, 2009), \emph{ggthemes}
  (Version 3.4.0; Arnold, 2017), \emph{kableExtra} (Version 0.7.0; Zhu,
  2018), \emph{keras} (Version 2.1.4; Allaire \& Chollet, 2018),
  \emph{knitr} (Version 1.20; Xie, 2015), \emph{leaflet} (Version 2.0.0;
  Cheng, Karambelkar, \& Xie, n.d.), \emph{padr} (Version 0.4.0; Thoen,
  2017), \emph{papaja} (Version 0.1.0.9709; Aust \& Barth, 2018),
  \emph{raster} (Version 2.6.7; Hijmans, 2017b), \emph{RColorBrewer}
  (Version 1.1.2; Neuwirth, 2014), \emph{readr} (Version 1.1.1; Wickham,
  Hester, \& Francois, 2017), \emph{rgdal} (Version 1.2.16; R. Bivand,
  Keitt, \& Rowlingson, 2017), \emph{scales} (Version 0.5.0; Wickham,
  2017), \emph{sp} (Version 1.2.7; Hijmans, 2017a; Pebesma \& Bivand,
  2005), \emph{tibbletime} (Version 0.1.0; Vaughan \& Dancho, 2018), and
  \emph{tidyr} (Version 0.8.0; Wickham \& Henry, 2018).}.

All the code is available on a public repository (Sobrado, 2018).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluating the performance of the map building and assigning
  functions.
\item
  Comparing the performance of PMMI compared to alternative methods
  {[}(Barnett \& Onnela, 2016; Palmius et al., 2017){]} and a baseline
  model using cross-validation.
\item
  Comparing the performance of PPMI compared to the aformentioned
  methods using objective ground-truth data (public transportation
  time-stamps locations).
\end{enumerate}

@@

Before we can evaluate the accuracy of the imputations, it is essential
to evaluate how well noise in the data has been smoothed. All methods
smoothen noise differently in an attempt to extract true location
\(G(t)\) from measurements \(D_t\). In the subsequent cross-validation
step we will remove random smoothened blocks and compare the imputed
value to the smoothened value. For this reason we want to ensure that
the smoothened method actually does reflect the true location \(G(t)\).
Otherwise we run the risk of over-estimating the accuracy of a model in
the sense that we will compute the extent to which an imputation method
can correctly impute a smoothened value which has little to do with the
true location \(G(t)\).

In order to evaluate the map building and binning we will first visually
evaluate the paths and pause locations. A visual evaluation of paths
superimposed on is an established way to heuristically check their
accuracy (e.g. Brunsdon, 2007). Then, let the average distance between
the actual measured point and the binned point be the \emph{deviation
distance} \(\delta_{\text{dev}}\). With respect to the deviation
distance \(\delta_{\text{dev}}\), we expect:

@@

\subsubsection{Imputation algorithm performance in
cross-validation.}\label{imputation-algorithm-performance-in-cross-validation.}

We will compare the performance of PMMI, Palmius et al. (2017) and
Barnett and Onnela (2016). In addition, we will also compute two
baseline models. The first one is a naive model, which simply imputes
the previous observed value. The second one we will call the
\enquote{home} model, which imputes that the individual is at home at
all missing timepoints. Using these two models as baselines makes sense
given the high degree of autocorrelation in the data and that
individuals tend to spend most of their time at home. To compare the
performance of the aforementioned methods we will remove 25\% of
measured time intervals at random within a four week period. We will
make our comparisons for intervals of 5 minutes, 1 hour and 1 day. In
other words, we will remove 25\% of time intervals at random, while
varying the duration of the time intervals removed.

@@

For the entire period examined period the we find a deviance of 40
meters and a median deviance of 15 meters. Around 69\% of the deviance
values are within their corresponding accuracy value, which is close to
the theoretical 67\% value that is expected. Approximately 9\% of values
were not taken into account when creating the bins,given that their the
accuracy \(a\) exceeded \(a_{\text{Path Lim}} = 300\).
